---
title: "Chi-square Test"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: github_document
---

### Cleaning and loading packages

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
remove(list = ls())
library(tidyverse)
```

# A F2 population

The $F_{2}$ population is derived from the selfing of the $F_{1}$ progeny (Aa). The $F_{2}$ population has three genotypes: $AA$, $Aa$, and $aa$, which segregate in a 1:2:1 ratio $(0.25, 0.5, 0.25)$.

### Loading dataset - maize

```{r}
maize_df <- read.csv("maize.csv")

head(maize_df)
dim(maize_df)
markers <- maize_df[,2:13]
```

### Markers info

So, using **ggplot** and **dplyr**, both packages from the **tidyverse**, we can organize and plot a nicer chart to verify the markers segregation distribution.

```{r}
markers %>% mutate(ind = rownames(markers)) %>%
  pivot_longer(-ind, names_to = "mrk", values_to = "value") %>% 
  ggplot(aes(x = ind, y = mrk, fill = as.factor(value)))+
  geom_tile()+
  scale_fill_manual(name = "Values",
                    values = c("0" = "red",
                                "1" = "blue",
                                "2" = "yellow"))+
  labs(x = "Individuals",
       y = "Markers")+
  theme_minimal()+
  theme(axis.title = element_text(size = 16, face = "bold"),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

### Chi-square tests ($\chi$²)

-   $H_0$: ratio of $aa$(0), $Aa$(1) and $AA$(2) is 1:2:1;
-   $H_a$: ratio of $aa$(0), $Aa$(1) and $AA$(2) is not 1:2:1;

```{r}
exp <- c(0.25, 0.5, 0.25)

p.value <- c()

M <- ncol(markers)

for (m in 1:M){
  obs <- table(markers[,m])
  p.value[m] <- chisq.test(x = obs, p = exp)$p.value
} 

print(p.value)
```

Plotting the distribution of p-values.

```{r}
p.value_df <- data.frame(marker = 1:length(p.value),
                         p.value = p.value)

p.value_df %>%
  mutate(marker = factor(marker)) %>%
  ggplot(aes(x = marker, y = p.value)) +
  geom_point() +
  geom_hline(yintercept = 0.05, color = "red", linetype = "dashed") +  
  geom_hline(yintercept = 0.05/M, color = "blue",  linetype = "dashed") +  
  labs(x = "Markers", y = "P-value") +
  theme_bw() +
  theme(axis.title = element_text(size = 16, face = "bold"),
        axis.text = element_text(size = 12))
```

Verifying how many values fall outside the limits.

```{r}
table(p.value > 0.05)
```

Applying Bonferroni correction:

```{r}
table(p.value > 0.05/M)
```

# A Backcross population

A backcross population is obtained by crossing two parents, one homozygous $P_1$ $(AA)$ and one heterozygous $P_2$ $(Aa)$, producing an $F_1$ $(Aa)$ progeny . This is followed by a cross between the $F_1$ and one of the original parents, known as the recurrent parent. Thus, the expected segregation ratio is 1:1 after a single backcross generation.

### Loading dataset - mouse

```{r}
remove(list = ls())

mouse_df <- read.csv("mouse.csv")

head(mouse_df)

dim(mouse_df)
```

### Makers info

So, using \*\*ggplot\*\$ **and** dplyr**, both packages from the** tidyverse\*\*, we can organize and plot a nicer chart to verify the markers segregation distribution.

```{r}
markers <- mouse_df[,2:15]

markers %>% mutate(ind = rownames(markers)) %>%
  pivot_longer(-ind, names_to = "mrk", values_to = "value") %>% 
  ggplot(aes(x = ind, y = mrk, fill = as.factor(value)))+
  geom_tile()+
  scale_fill_manual(name = "Values",
                    values = c("0" = "red",
                               "1" = "blue",
                               "2" = "yellow"))+
  labs(x = "Individuals",
       y = "Markers")+
  theme_minimal()+
  theme(axis.title = element_text(size = 16, face = "bold"),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

### Chi-square Test ($\chi$²)

-   $H_0$: ratio of $aa$(0),and $AA$(1) is 1:1;
-   $H_a$: ratio of $aa$(0),and $AA$(1) is not 1:1;

```{r}
exp <- c(0.5, 0.5)

p.value <- c()

M <- ncol(markers)

for (m in 1:M){
  obs <- table(markers[,m])
  p.value[m] <- chisq.test(x = obs, p = exp)$p.value
} 

print(p.value)
```

Plotting the distribution of p-values.

```{r}
p.value_df <- data.frame(marker = 1:length(p.value),
                         p.value = p.value)

p.value_df %>% mutate(marker = factor(marker)) %>%
  ggplot(aes(x = marker, y = p.value))+
  geom_point()+
  geom_hline(yintercept = 0.05, color = "red", linetype = "dashed")+
  geom_hline(yintercept = 0.05/M, color = "blue", linetype = "dashed")+
  labs(x = "Markers", y = "P-value")+
  theme_bw()+
  theme(axis.title = element_text(size = 16, face = "bold"),
        axis.text = element_text(size = 12))
```

Verifying how many values fall outside the limits,

```{r}
table(p.value > 0.05)
```

Applying Bonferroni correction:

```{r}
table(p.value > 0.05/M)
```


